Headgroup structure and cation binding in phosphatidylserine lipid bilayers

### 2019-05-22 Manuscript/manuscriptPS.tex
Updates based on comments and text from Tom Piggot sent by email

\author{Hanne Antila}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{NMR group - Institut for Physics, Martin-Luther University Halle-Wittenberg}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Markus S. Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot}
\affiliation{Southampton, United Kingdom}
%\author{NMRlipids collaboration}
%\affiliation{nmrlipids.blogspot.fi}

@@

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eukaryotic membranes. In red blood cells, for example,
PS lipids compose 8.5\% of the total lipid weight. 
The abundance, however, varies between different organelles, and up to
25-35\% of the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14} consists of PS lipids.
PS lipids are vastly important biomolecules that interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some protein domains interact specifically with PS lipids,
while other protein sites attract PS lipids by nonspecific electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the deciphering structural details
of lipid headgroups and the details of cation binding
are crucial for understanding the PS-mediated processes in the cell membranes.

Experimental studies have indicated that the
PS headgroup is more rigid than the phosphatidylcholine (PC)
owing to electrostatic interactions or the formation of hydrogen bonding network between the headgroups~\cite{browning80,buldt81}.
While most monovalent ions interact weakly with
PS-containing bilayers, multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
The dehydrated complexes of PS headgroup with calcium ions can even lead to
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}. Mixing PS lipids with PC lipids reduces their propensity to form strong complexes with multivalent ions and makes the PS headgroup less rigid~\cite{browning80,buldt81,roux90,roux91}.
That said, some studies suggest Ca$^{2+}$ has similar specific binding affinity 
to negatively charged and zwiterionic phospholipids, and that
the increased cation binding to negatively charged lipid bilayers arises only due
to the increased local cation concentration in the membrane vicinity~\cite{seelig90,sinn06}.

@@

@@

lipid/counter-ions  & force field for lipids / ions  & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline

@@

POPS/Na$^{+}$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36uaPOPS298K}  & \tabularnewline

@@

POPS/Na$^{+}$  & lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \cite{lipid17POPSjcions}  & \tabularnewline
POPS/Na$^{+}$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \cite{lipid17POPSff99ions}  & \tabularnewline
POPS/Na$^{+}$  & Berger \cite{mukhopadhyay04,??}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{bergerPOPS298K}  & \tabularnewline
POPS/Na$^{+}$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{ckp1POPS303K}  & \tabularnewline
POPS/Na$^{+}$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{ckp2POPS303K}  & \tabularnewline

@@

DOPS/Na$^{+}$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{charmm36uaDOPS303K}  & \tabularnewline
DOPS/Na$^{+}$  & lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \cite{lipid17DOPSjcions}  & \tabularnewline
DOPS/Na$^{+}$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \cite{lipid17DOPSff99ions}  & \tabularnewline
DOPS/Na$^{+}$  & Berger \cite{mukhopadhyay04,??}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{bergerDOPS303K}  & \tabularnewline
DOPS/Na$^{+}$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{ckp1DOPS303K}  & \tabularnewline
DOPS/Na$^{+}$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{ckp2DOPS303K}  & \tabularnewline

@@

We have recently demonstrated that the lipid C--H bond order parameters, $S_\mathrm{CH}$,
can be utilized to resolve such controversies~\cite{botan15,catte16}. The $S_\mathrm{CH}$ can be
measured from NMR experiments with high accuracy and directly compared to simulations
in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}. Using this approach,
it established that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, and that the cation binding to PC
lipid bilayers is overestimated \cite{catte16}.
%
%Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
%\todo{Should we leave the mention of ECC out from the Introduction
%(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}

@@

POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP \cite{??}  & 0  & 110 & ?  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{POPCpopsGROMOSCKPwithNa} \tabularnewline
POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM \cite{??}  & 0  & 110 & ?  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{POPCpopsGROMOSCKPMwithNa} \tabularnewline

@@

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were calculated by first averaging over time separately
for each lipid in the system, and then calculating the average and
the standard error of the mean over the different lipids. The analysis was conducted using a
Python program ({\tt calcOrderParameters.py}, available in Ref. \citenum{MATCHgit}) that utilizes the
MDAnalysis library \cite{agrawal11,gowers16}. 
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

The different PS MD models produce highly varied  headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the SI ),
as previously observed also for PC lipids \cite{botan15}, and none of the models produces a set of order parameters in full agreement with the experiments.
The models perform generally less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
which complicates the interpretation of structural differences between PC and PS headgroups. However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36ua) do replicate the larger-than-in-PC
forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
also observed in the experiments (Fig. \ref{HGorderParameters}).

\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPSformatted.pdf}
  \caption{\label{NAdensPOPS}
    Counterion density profiles along the membrane normal and areas per lipid (A$_{\rm pl}$)
    of POPS lipid bilayer from simulations with different force fields.
    The experimental area per lipid is~62.7~$\pm$~1.3~\AA$^2$~\cite{pan14}.
  }
  \todo{Commented by M. Javanainen in blog:
    MacRog pure POPS is simulated with Verlet cutoff scheme, Piggot is rerunning with group cutoff scheme. Check if affects results \& update figures when ready} \\
\end{figure}

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
    Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\citenum{ferreira16}.
    The value from the CHARMM36 simulation with sodium is the average of two independent simulations and
    the error bar is given as the difference between the results divided by two.
  }
  \todo{After we know which force field is used for POPC in Gromos-CKP simulations, we might be able to
    add Gromos-CKP data into this plot.}
\end{figure}

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
  \caption{\label{HGorderparametersPSvsPC}
    Modulation of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
    Experimental values with the signs are measured for pure POPS system in this work.
    The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
    Because the experimental values of POPS in pure and mixed bilayers come from $^{13}$C NMR (this work) and $^2$H NMR (Ref. \citenum{roux90}), respectively,
    the error bars of 0.02 are used here \cite{botan15,ollila16}.
    The y-axis for the $\alpha$-carbon results of POPS (bottom) is shifted
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
    The values from CHARMM36 and Gromos simulations are averages of two independent simulations and
    the error bars are given as the difference between the results divided by two.
  }
\end{figure}

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers compared to PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. We explore this by quantifying the counterion density profiles along the membrane normal accompanied by the areas per lipid in Fig.~\ref{NAdensPOPS}.
The force fields studied show significant differences in both binding affinity
and distribution of ions at the interface.
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in Gromos-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller values.
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36ua simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
The results are in line with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.

@@

\subsection{Ca$^{2+}$ binding affinity to bilayers with negatively charged PS lipids}
\begin{figure}[tb]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixtureLOWconsformatted.pdf}
  \caption{\label{CAdensPCPSmixture}
    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields.
    The smallest simulated CaCl$_2$ concentrations are shown.
    The density profiles for all the simulated concentrations are given in SI figure \ref{CAdensPCPSmixtureALL}.
  }
  \todo{Should we include also counterions into the plot?} \\
\end{figure}

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
  OHSO acknowledges financial support from Academy of Finland (315596),
  Integrated Structural Biology Research Infrastructure of
  Helsinki Institute of Life Science (Instruct-HiLIFE), and
  CSC-IT center for science for computational resources.
  MJ acknowledges financial support from the Emil Aaltonen foundation and
  and CSC-IT center for science for computational resources.
\end{acknowledgments}

>>>>

\author{Hanne Antila}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{NMR group - Institut for Physics, Martin-Luther University Halle-Wittenberg}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Markus S. Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas J. Piggot}
\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}
%\author{NMRlipids collaboration}
%\affiliation{nmrlipids.blogspot.fi}

@@

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eukaryotic membranes. In red blood cells, for example,
PS lipids compose 8.5\% of the total lipid weight. 
The abundance, however, varies between different organelles, and up to
25-35\% of the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14} consists of PS lipids.
PS lipids are vastly important biomolecules that interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some protein domains interact specifically with PS lipids,
while other protein sites attract PS lipids by nonspecific electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, deciphering the structural details
of lipid headgroups and the details of cation binding
are crucial for understanding the PS-mediated processes in the cell membranes.

Experimental studies have indicated that the
PS headgroup is more rigid than the phosphatidylcholine (PC)
owing to electrostatic interactions or the formation of hydrogen bonding network between the headgroups~\cite{browning80,buldt81}.
While most monovalent ions interact weakly with
PS-containing bilayers, multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
The dehydrated complexes of PS headgroups with calcium ions can even lead to
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}. Mixing PS lipids with PC lipids reduces their propensity to form strong complexes with multivalent ions and makes the PS headgroup less rigid~\cite{browning80,buldt81,roux90,roux91}.
That said, some studies suggest Ca$^{2+}$ has similar specific binding affinity 
to negatively charged and zwiterionic phospholipids, and that
the increased cation binding to negatively charged lipid bilayers arises only due
to the increased local cation concentration in the membrane vicinity~\cite{seelig90,sinn06}.

@@

  $^*$ Force field parameters for PS lipids generated for this work, for full details see the supplementary information.

@@

lipid/counter-ions  & force field for lipids & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline

@@

POPS/Na$^{+}$  & CHARMM36-UA$^*$ \cite{venable13,lee14} & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{charmm36uaPOPS298K}  & \tabularnewline

@@

POPS/Na$^{+}$  & Lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \cite{lipid17POPSjcions}  & \tabularnewline
POPS/Na$^{+}$  & Lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 298  & 2$\times$600  & 2$\times$100  & \cite{lipid17POPSff99ions}  & \tabularnewline
POPS/Na$^{+}$  & Berger \cite{mukhopadhyay04}  & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{bergerPOPS298K}  & \tabularnewline
POPS/Na$^{+}$  & GROMOS-CKPM \cite{Chandrasekhar03,kukol09,piggot12} & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{ckp1POPS303K}  & \tabularnewline
POPS/Na$^{+}$  & GROMOS-CKP \cite{Chandrasekhar03,kukol09,piggot12} & 128  & 4480  & 298  & 2$\times$500  & 2$\times$100  & \cite{ckp2POPS303K}  & \tabularnewline

@@

DOPS/Na$^{+}$  & CHARMM36-UA$^*$ \cite{venable13,lee14} & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{charmm36uaDOPS303K}  & \tabularnewline
DOPS/Na$^{+}$  & Lipid17 \cite{gould18} / JC \cite{joung08}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \cite{lipid17DOPSjcions}  & \tabularnewline
DOPS/Na$^{+}$  & Lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128  & 4480  & 303  & 2$\times$600  & 2$\times$100  & \cite{lipid17DOPSff99ions}  & \tabularnewline
DOPS/Na$^{+}$  & Berger \cite{mukhopadhyay04}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{bergerDOPS303K}  & \tabularnewline
DOPS/Na$^{+}$  & GROMOS-CKPM$^*$ \cite{piggot12} & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{ckp1DOPS303K}  & \tabularnewline
DOPS/Na$^{+}$  & GROMOS-CKP$^*$ \cite{piggot12}  & 128  & 4480  & 303  & 2$\times$500  & 2$\times$100  & \cite{ckp2DOPS303K}  & \tabularnewline

@@

We have recently demonstrated that the lipid C--H bond order parameters, $S_\mathrm{CH}$,
can be utilized to resolve such controversies~\cite{botan15,catte16}. The $S_\mathrm{CH}$ can be
measured from NMR experiments with high accuracy and directly compared to simulations
in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}. Using this approach,
it has been established that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, that the cation binding to PC
lipid bilayers is overestimated~\cite{catte16}, and that the splitting of the order parameters at
carbon 2 in {\it sn}-2 lipid tails is correctly reproduced only in the CHARMM36 force field~\cite{piggot17}.% [https://pubs.acs.org/doi/abs/10.1021/acs.jctc.7b00643].
%
%Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
%\todo{Should we leave the mention of ECC out from the Introduction
%(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}

@@

POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKP$^*$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{POPCpopsGROMOSCKPwithNa} \tabularnewline
POPC:POPS (5:1)/Na$^{+}$  & GROMOS-CKPM$^*$ \cite{piggot12}  & 0  & 110 & 4620  & 0  & 298  & 2$\times$500  & 2$\times$100  & \cite{POPCpopsGROMOSCKPMwithNa} \tabularnewline

@@

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were calculated by first averaging over time separately
for each lipid in the system, and then calculating the average and
the standard error of the mean over the different lipids. The analysis can be conducted using a
Python program ({\tt calcOrderParameters.py}, available in Ref. \citenum{MATCHgit}) that utilizes the
MDAnalysis library \cite{agrawal11,gowers16}.
For united atom models, the positions of hydrogens were generated before the order parameter calculation using {\tt protonate} tool
of the Gromacs 3 sofware package \cite{gromacsMANUAL}.
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

The different PS MD models produce highly varied  headgroup and glycerol backbone order parameters (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS} and section~\ref{Diheds} in the SI ),
as previously observed also for PC lipids \cite{botan15}, and none of the models produces a set of order parameters in full agreement with the experiments.
The models perform generally less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
which complicates the interpretation of structural differences between PC and PS headgroups. However, concentrating on the headgroup, we see that the best performing models (Slipids, CHARMM36 and CHARMM36-UA) do replicate the larger-than-in-PC
forking of the $\alpha$-carbon observed in the experiments and that the Slipids force field additionally correctly produces the significantly
smaller $\beta$-carbon order parameter for PS compared to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15})
also observed in the experiments (Fig. \ref{HGorderParameters}).

\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPSformatted.pdf}
  \caption{\label{NAdensPOPS}
    Counterion density profiles along the membrane normal and areas per lipid (A$_{\rm pl}$)
    of POPS lipid bilayer from simulations with different force fields.
    The vertical green bars indicate the location of the phosphate density peak.
    The experimental area per lipid is~62.7~$\pm$~1.3~\AA$^2$~\cite{pan14}.
  }
%  \todo{Commented by M. Javanainen in blog:
%    MacRog pure POPS is simulated with Verlet cutoff scheme, Piggot is rerunning with group cutoff scheme. Check if affects results \& update figures when ready} \\
\end{figure}

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
    Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\citenum{ferreira16}.
    The value from the CHARMM36 simulation with sodium is the average of two independent simulations and
    the error bar is given as the difference between the results divided by two.
  }
  \todo{After we know which force field is used for POPC in GROMOS-CKP simulations, we might be able to
    add GROMOS-CKP data into this plot.}
\end{figure}

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
  \caption{\label{HGorderparametersPSvsPC}
    Modulation of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
    Experimental values with the signs are measured for pure POPS system in this work.
    The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
    Because the experimental values of POPS in pure and mixed bilayers come from $^{13}$C NMR (this work) and $^2$H NMR (Ref. \citenum{roux90}), respectively,
    the error bars of 0.02 are used here \cite{botan15,ollila16}.
    The y-axis for the $\alpha$-carbon results of POPS (bottom) is shifted
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
    The values from CHARMM36 and GROMOS simulations are averages of two independent simulations and
    the error bars are given as the difference between the results divided by two.
  }
\end{figure}

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers compared to PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups. We explore this by quantifying the counterion density profiles along the membrane normal accompanied by the areas per lipid in Fig.~\ref{NAdensPOPS}.
The force fields studied show significant differences in both binding affinity
and distribution of ions at the interface.
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in GROMOS-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller values (Fig.~\ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36-UA simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
The results are in line with the previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.

@@

\subsection{Ca$^{2+}$ binding affinity to bilayers with negatively charged PS lipids}
\begin{figure}[tb]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixtureLOWconsformatted.pdf}
  \caption{\label{CAdensPCPSmixture}
    Number density profiles of Ca$^{2+}$ (solid line) and Cl$^-$ (dashed line) from POPC:POPS (5:1) mixtures
    simulated with different force fields. The vertical green bars indicate the location of the phosphate density peak.
    The smallest simulated CaCl$_2$ concentrations are shown.
    The density profiles for all the simulated concentrations are given in SI figure \ref{CAdensPCPSmixtureALL}.
  }
\end{figure}

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
  MJ acknowledges financial support from the Emil Aaltonen foundation and
  and CSC-IT center for science for computational resources.
  OHSO acknowledges financial support from Academy of Finland (315596),
  Integrated Structural Biology Research Infrastructure of
  Helsinki Institute of Life Science (Instruct-HiLIFE), and
  CSC-IT center for science for computational resources.
  TJP wishes to acknowledge the use of the Iridis computing resources at the University of Southampton
\end{acknowledgments}
